---
title: "Introduction to santoku"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{santoku}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


## Installation

You can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("hughjonesd/santoku")
```

## Basic usage

Use `chop()` like `cut()` to cut your data up:

```{r}
library(santoku)
x <- runif(10, 0, 10)
(chopped <- chop(x, breaks = 0:10))
data.frame(x, chopped)
```

`chop()` returns a factor.

If data is beyond the limits of `breaks`, they will be extended automatically:

```{r}
chopped <- chop(x, breaks = 3:7)
data.frame(x, chopped)
```

To chop a single number into a separate category, put the number twice in
`breaks`:

```{r}
x_fives <- x
x_fives[1:5] <- 5
chopped <- chop(x_fives, c(2, 5, 5, 8))
data.frame(x_fives, chopped)
```


To quickly produce a table of chopped data, use `tab()`:

```{r}
tab(x, c(2, 5, 8))
```

## More ways to chop

To chop into fixed-width intervals, starting at the minimum value, use
`chop_width()`:

```{r}
chopped <- chop_width(x, 2)
data.frame(x, chopped)
```

To chop into groups with a fixed *size* (i.e. number of members), use `chop_size()`:

```{r}
chopped <- chop_size(x, 4)
table(chopped)
```


To chop into `n` equal sized groups use `chop_equal()`:

```{r}
chopped <- chop_equal(x, 3)
table(chopped)
```


To chop data up by quantiles, use `chop_quantiles()`:

```{r}
chopped <- chop_quantiles(x, c(0.25, 0.5, 0.75))
data.frame(x, chopped)
```

To chop data by standard deviations around the mean, use `chop_mean_sd()`:

```{r}
chopped <- chop_mean_sd(x)
data.frame(x, chopped)
```

`tab_size()`, `tab_width()` and `tab_mean_sd()` act similarly to `tab()`,
calling the related `chop_` function and then `table()`.

```{r}
tab_width(x, 2)
tab_size(x, 4)
tab_mean_sd(x)
```

# Advanced usage

You can change factor labels with the `labels` argument:

```{r}
chopped <- chop(x, c(2, 5, 8), labels = c("Lowest", "Low", "Higher", "Highest"))
data.frame(x, chopped)
```

You need one more label than there are breaks.

To label intervals with a dash, use `lbl_dash()`:

```{r}
chopped <- chop(x, c(2, 5, 8), lbl_dash())
data.frame(x, chopped)
```

For arbitrary formatting use `lbl_format()` and `sprintf`-style format strings:

```{r}
chopped <- chop(x, c(2, 5, 8), lbl_format("%s to %s"))
data.frame(x, chopped)
```

To number intervals in order use `lbl_numerals()`:

```{r}
chopped <- chop(x, c(2, 5, 8), lbl_numerals())
data.frame(x, chopped)
```

You can use letters or even roman numerals:

```{r}
chop(x, c(2, 5, 8), lbl_letters())
chop(x, c(2, 5, 8), lbl_roman())
```

By default, `chop()` extends `breaks` if necessary. If you don't want
that, set `extend = FALSE`:

```{r}
chopped <- chop(x, c(3, 5, 7), extend = FALSE)
data.frame(x, chopped)
```

Data outside the range of `breaks` will become `NA`.

By default, intervals are closed on the left, i.e. they include their left
endpoints. If you want right-closed intervals, use `brk_right()`:

```{r}
y <- 1:10
data.frame(
        y = y, 
        left_closed = chop(y, 2:8), 
        right_closed = chop(y, brk_right(2:8))
      )
```

The last finite interval is right-closed (or if you use `brk_right`, the first finite interval is left-closed). If you don't want that, use `brk_left()` explicitly and set `close_end = FALSE`:

```{r}
z <- 1:5
data.frame(
  z = z,
  rightmost_closed = chop(1:5, brk_left(1:5)),
  rightmost_open   = chop(1:5, brk_left(1:5, close_end = FALSE))
)

```


If you want to chop repeatedly with the same arguments, create your own
`knife`:

```{r}
chop_by_quartiles <- knife(
        breaks = brk_quantiles(c(0.25, 0.5, 0.75)), 
        labels = lbl_dash()
      )

chop_by_quartiles(x)
table(chop_by_quartiles(rnorm(50)))

```

## Speed

The core of santoku is written in C++. It is reasonably fast:

```{r}
microbenchmark::microbenchmark(
        chop(rnorm(1e5), -2:2),
         cut(rnorm(1e5), -2:2)
      )
```
