---
title: "Introduction to santoku"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to santoku}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
set.seed(23479)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


## Installation

To install santoku, run:

``` r
install.packages("santoku")
```

## Basic usage

Use `chop()` like `cut()` to cut your data up:

```{r}
library(santoku)
x <- runif(10, 0, 10)
(chopped <- chop(x, breaks = 0:10))
data.frame(x, chopped)
```

`chop()` returns a factor.

If data is beyond the limits of `breaks`, they will be extended automatically:

```{r}
chopped <- chop(x, breaks = 3:7)
data.frame(x, chopped)
```

To chop a single number into a separate category, put the number twice in
`breaks`:

```{r}
x_fives <- x
x_fives[1:5] <- 5
chopped <- chop(x_fives, c(2, 5, 5, 8))
data.frame(x_fives, chopped)
```


To quickly produce a table of chopped data, use `tab()`:

```{r}
tab(1:10, c(2, 5, 8))
```

## More ways to chop

To chop into fixed-width intervals, starting at the minimum value, use
`chop_width()`:

```{r}
chopped <- chop_width(x, 2)
data.frame(x, chopped)
```

To chop into exactly `intervals` fixed-with intervals, use `chop_evenly()`:

```{r}
chopped <- chop_evenly(x, intervals = 3)
data.frame(x, chopped)
```

To chop into groups with a fixed number of members, use `chop_n()`:

```{r}
chopped <- chop_n(x, 4)
table(chopped)
```


To chop into a fixed number of equal-sized groups, use `chop_equally()`:

```{r}
chopped <- chop_equally(x, groups = 5)
table(chopped)
```


To chop data up by quantiles, use `chop_quantiles()`:

```{r}
chopped <- chop_quantiles(x, c(0.25, 0.5, 0.75))
data.frame(x, chopped)
```

To chop data by standard deviations around the mean, use `chop_mean_sd()`:

```{r}
chopped <- chop_mean_sd(x)
data.frame(x, chopped)
```

`tab_n()`, `tab_width()`, `tab_evenly()` and `tab_mean_sd()` act similarly to
`tab()`, calling the related `chop_` function and then `table()`.

```{r}
tab_n(x, 4)
tab_width(x, 2)
tab_evenly(x, 5)
tab_mean_sd(x)
```


You can chop dates too:

```{r}
library(lubridate)
y2k <- as.Date("2000-01-01") + 0:365
months <- chop_width(y2k, months(1))
table(months)
```



# Advanced usage

You can change factor labels with the `labels` argument:

```{r}
chopped <- chop(x, c(2, 5, 8), labels = c("Lowest", "Low", "Higher", "Highest"))
data.frame(x, chopped)
```

You need as many labels as there are intervals - one fewer than `length(breaks)`
if your data doesn't extend beyond `breaks`, one more than `length(breaks)` if
it does.

To label intervals with a dash, use `lbl_dash()`:

```{r}
chopped <- chop(x, c(2, 5, 8), lbl_dash())
data.frame(x, chopped)
```

To label intervals in order use `lbl_seq()`:

```{r}
chopped <- chop(x, c(2, 5, 8), lbl_seq())
data.frame(x, chopped)
```

You can use numerals or even roman numerals:

```{r}
chop(x, c(2, 5, 8), lbl_seq("(1)"))
chop(x, c(2, 5, 8), lbl_seq("i."))
```

For arbitrary formatting use `lbl_format()` and `sprintf`-style format strings:

```{r}
chopped <- chop(x, c(2, 5, 8), lbl_format("%s to %s"))
data.frame(x, chopped)
```

Or use a formatter from the `{scales}` package:

```{r}
library(scales)
r <- runif(10)
chopped <- chop(r, c(.3, .5, .7), lbl_intervals(fmt = label_percent(0.1)))
data.frame(r, chopped)
```

By default, `chop()` extends `breaks` if necessary. If you don't want
that, set `extend = FALSE`:

```{r}
chopped <- chop(x, c(3, 5, 7), extend = FALSE)
data.frame(x, chopped)
```

Data outside the range of `breaks` will become `NA`.

By default, intervals are closed on the left, i.e. they include their left
endpoints. If you want right-closed intervals, set `left = FALSE`:

```{r}
y <- 1:5
data.frame(
        y = y, 
        left_closed = chop(y, 1:5), 
        right_closed = chop(y, 1:5, left = FALSE)
      )
```

If you want to close off the last interval, set `close_end = TRUE`:

```{r}
data.frame(
  y = y,
  rightmost_open = chop(y, 1:5),
  rightmost_closed   = chop(y, 1:5, close_end = TRUE)
)

```

